Supplemental Material: Multidimensional analysis and detection of informative features in human brain white matter

### 2021-04-20 reports/manuscript/paper_changes.tex
Add tractography details

The present work aims to balance predictive accuracy with descriptive power
\cite{Murdoch2019-ax, Breiman2001-uz} by capitalizing on all of the available
data across all bundles, while also retaining and elucidating spatial
information about the locations that are most informative for discriminative
performance.
\added{Unlike many previous analyses of white matter (e.g., using TBSS \cite{smith2006tract} and connectometry \cite{yeh2016connectometry}), which focus on classical inference of the differences between groups or individuals in terms of their white matter properties, the focus of the present work is on the combination of features that facilitate accurate prediction of the individual differences in a \emph{phenotype}}.
\added{For example, an accurate prediction of which individuals in a group have a particular disease, or a prediction of their age with small error. The distinct goals of \emph{inference} and \emph{prediction} are not necessarily incongruent, but can sometimes be in tension \cite{Bzdok2020-sd}. This distinction is also similar to the distinction between ``encoding'' and ``decoding'' used in the functional MRI literature \cite{Naselaris2011-ss}. In the present work}, we predict the phenotypical variance in a group of subjects (or classify group membership), based on a linear combination of the features estimated with \replaced{tract profiling}{tractometry}.

@@

The $\beta$ coefficients exhibit high bundle level sparsity; only some
bundles are important, which can be be confirmed by observing the value of
$\alpha$, the regularization hyperparameter that controls the mixture of the
Group Lasso and lasso penalties, selected through nested cross-validation
(see \cref{eq:sgl}).
If $\alpha$ is closer to zero, it indicates that the phenotype in question
preferentially correlates with only a few groups of covariates. For the ALS
dataset, \added{the SGL model has} $\alpha = $ \alsLRatio
~\added{and the PCR-SGL model has \protect$\alpha = $ \protect\alsLRatioGpca},
confirming that the white matter correlates of ALS reside mostly in one
bundle, namely the CST.

@@

To test the performance of SGL \deleted{with tractometry data} in a continuous
regression task, we focus here on the prediction of biological age in three
datasets named WH, HBN, and Cam-CAN (see the Methods section for a description of each dataset). Prediction of ``brain
age'' is a commonly undertaken task in neuroimaging machine learning, in part
because these predictions, and deviations therefrom, may be diagnostic of
overall brain health (for a review, see Cole et al.~\cite{Cole2019-rz}).
However, as Nelson et al.~\cite{nelson2019biomarkers} have observed, aging
biomarkers are subject to unique challenges and tend to be noncausative. Our
interest in aging here relies on its utility as a methodological benchmark.
Biological age operates on a natural scale, with meaningful and easily
understood units, and it's popularity as a machine learning target makes it
valuable for comparisons with other studies.

@@

To evaluate the fit of the model, we used a nested cross-validation
procedure. In this procedure, batches of subjects are held out. For each
batch (or fold), the model is fully fit without this data. Then, once the
parameters are fixed, the model is applied to predict the ages of held out
subjects based on the linear coeffiecients. This scheme automatically finds
the right level of sparsity and fits the coefficients to the ill-posed linear
model, while guarding against overfitting. SGL accurately predicts the age of
the subjects in this procedure, with a median absolute error of {\whMae},
\added{{\hbnMae}}, and {\camcanMae} years for the WH, HBN, and Cam-CAN datasets,
respectively and coefficients of determination $R^2 = $ {\whRsq} , \added{{\hbnRsq}}, and {\camcanRsq},
respectively (see \cref{fig:age-results}, top panels). The
predictions for Cam-CAN are competitive with a recent state of the art
prediction \cite{mcpherson2020single}, which used streamline density to
estimate the brain's structural connectivity and achieved $R^2 = 0.63$. The
median absolute errors are also lower than the results of a recent study that
predicted age in a large sample that included the Cam-CAN data, and was based
on diffusion MRI features \cite{Richard2018-ux}.

@@

Specifically, we demonstrated that the algorithm correctly detects the fact
that ALS, which is a disease of lower motor neurons, is localized to the
cortico-spinal tract. This recapitulates the results of previous analysis of
these same data, using a bundle-of-interest approach
\cite{sarica2017corticospinal}.
\added{As with the original study, it is unclear whether our strong predictive performance based on image properties that would be visible in patient scans or whether it is identifying a new subclinical disease biomarker.}
In contrast, for the analysis of biological
age, the coefficients identified by the algorithm are very widely distributed
across many parts of the white matter, mirroring previous results that show a
large and continuous distribution of life-span changes in white matter
properties \cite{yeatman2014lifespan}.
\added{We present age quintile bundle profiles and SGL \protect$\beta$ coefficients for all four datasets in the supplementary material. For age regression, substantial differences between the datasets preclude generalization across all three:}
\begin{enumerate*}[%
    label=(\roman*),%
    before=\unskip{: },%
    itemjoin={{, }},%
    itemjoin*={{, and }}]
    \item \added{different acquisition parameters, which can be challenging to harmonize \mbox{\cite{pinto2020harmonization}}}
    \item \added{different diffusion models, with DTI for the WH dataset and DKI for the HBN and Cam-CAN datasets}
    \item \added{different age ranges and distributions (which is evident in the figure legends for \cref{fig:age-results}), with HBN being a developmental dataset, while WH and Cam-CAN are lifespan maturation datasets}
    \item \added{different anatomical extents, with the WH streamlines truncated to remain with the bundle's bounding regions of interest (the default behavior in the legacy \texttt{mAFQ}) and the HBN and Cam-CAN streamlines allowed to retain their full extent from whole-brain tractography (the default behavior in \texttt{pyAFQ}).}
\end{enumerate*}
\added{In a biological characteristic that has widespread effects in the brain, validity is difficult to assess \mbox{\cite{lermausabiaga2019replication}}, so it is both unsurprising and vexing that we can do so well at predicting within each dataset, and yet do poorly in interpreting informative features across instruments.}

@@

\added{Another limitation of our approach is in its computational execution time,}
\added{which, in our experience, is roughly five times slower than simpler models}
\added{like the lasso. This slow-down is likely to be insignificant for ``production''}
\added{cycles in which researchers report their findings, but may be prohibitive}
\added{for ``development'' cycles in which researchers repeatedly train models while}
\added{adjusting their analysis pipeline. In practice, we circumvent this limitation}
\added{by developing our analysis infrastructure with fast, simple models, and substituting}
\added{slower, more performant models afterwards. The consistent \texttt{Scikit Learn}-based application programming interface of \texttt{AFQ-Insight} provides a good basis for this pattern.}

@@

The method is packaged as open-source software called AFQ-Insight that is
openly available at \url{https://github.com/richford/AFQ-Insight}, and
provides a clear API to allow for extensions of the method. The sofware
integrates within a broader automated fiber quantification software
ecosystem: AFQ \cite{yeatman2012tract} and pyAFQ \cite{kruper2021evaluating}, which extract
\replaced{tract profile}{tractometry} data from raw and processed dMRI datasets, as well as
AFQ-Browser, which visualizes \replaced{tract profiles}{tractometry} data and facilitates sharing of the
results of dMRI studies \cite{yeatman2018browser}. To facilitate
reproducibility and ease use of the software, the results presented in this
paper are also provided in
\url{https://github.com/richford/afq-insight-paper} as a series of Jupyter
notebooks \cite{kluyver2016jupyter}.

@@

@@

\item Diffusion MRI data from the Cambridge Centre for Ageing and
Neuroscience (Cam-CAN) ``CC700'' dataset
\cite{shafto2014cambridge,taylor2017cambridge}, containing data from 640
subjects with ages 18-88. These data were measured on a 3T Siemens TIM Trio
system and written informed consent was obtained from each participant. Voxel
resolution was \num{2x2x2}$\text{mm}^3$ with 30 non-colinear directions
measured for each of $b=1000$ $\text{s} / \text{mm}^2$ and $b=2000$
$\text{s} / \text{mm}^2$. The diffusion weighted images were acquired
with a twice refocused spin-echo sequence and the same preprocessing pipeline
used for the HBN dataset was applied to this data. We will refer to this
dataset as Cam-CAN.

@@

Data from the ALS and WH studies was processed in a similar manner,
using the Matlab Automated Fiber Quantification toolbox (\texttt{mAFQ})
\cite{yeatman2012tract}: streamlines representing fascicles of white
matter tracts were generated using a determinstic tractography algorithm
that follows the prinicpal diffusion direction of the diffusion tensor
in each voxel (STT) \cite{basser2000vivo}. \replaced{Eighteen major}{Major}
tracts\added{, which are enumerated in the supplemental material,} were identified
using multiple criteria: streamlines are selected as candidates for
inclusion in a bundle of streamlines that represents a tract if they
pass through known inclusion ROIs and do not pass through exclusion
ROIs \cite{wakana2007reproducibility}. In addition, a probabilistic
atlas is used to exclude streamlines which are unlikely to be part of
a tract \cite{Hua2008-sh}. Each streamline is resampled to 100 nodes
and the robust mean at each location is calculated by estimating the 3D
covariance of the location of each node and excluding streamlines that
are more than 5 standard deviations from the mean location in any node.
Finally, a bundle profile of tissue properties in each bundle was created
by interpolating the value of MRI maps of these tissue properties to the
location of the nodes of the resampled streamlines designated to each
bundle. In each of 100 nodes, the values are summed across streamlines,
weighting the contribution of each streamline by the inverse of the
\replaced{mahalanobis}{mahalnobis} distance of the node from the average of
that node across streamlines. This means that streamlines that are more
representative of the tract contribute more to the bundle profile, relative
to streamlines that are on the edge of the tract.

Data from the HBN and Cam-CAN studies were processed using the updated Python
Automated Fiber Quantification toolbox (\texttt{pyAFQ}; \cite{kruper2021evaluating}). In
addition to demonstrating the our analysis pipeline is robust to changes in
tractometry software, the use of the updated \texttt{pyAFQ} capitalized upon
the following improvements over the legacy Matlab version
\begin{enumerate*}[%
    label=(\roman*),%
    before=\unskip{: },%
    itemjoin={{, }},%
    itemjoin*={{, and }}]
    \item the ability to ingest data provided in the BIDS format
    \cite{gorgolewski2016brain}
    \item the calculation of diffusion kurtosis imaging (DKI
    \cite{jensen2005diffusion}) metrics
\end{enumerate*}
We will refer to the \texttt{mAFQ} and \texttt{pyAFQ} pipeline collectively
as \texttt{AFQ}.

>>>>

The present work aims to balance predictive accuracy with descriptive power
\cite{Murdoch2019-ax, Breiman2001-uz} by capitalizing on all of the available
data across all bundles, while also retaining and elucidating spatial
information about the locations that are most informative for discriminative
performance. To do so, we use a linear modeling approach, which aims to
redict phenotypical variance in a group of subjects (or classify group
membership), based on a linear combination of the features estimated with
\replaced{tract profiling}{tractometry}.

@@

The $\beta$ coefficients exhibit high bundle level sparsity; only some
bundles are important, which can be be confirmed by observing the value of
$\alpha$, the regularization hyperparameter that controls the mixture of the
Group Lasso and lasso penalties, selected through nested cross-validation
(see \cref{eq:sgl}).
If $\alpha$ is closer to zero, it indicates that the phenotype in question
preferentially correlates with only a few groups of covariates. For the ALS
dataset, \added{the SGL model has} $\alpha = $ \alsLRatio 
\added{and the PCR-SGL model has \protect$\alpha = $ \protect\alsLRatioGpca},
confirming that the white matter correlates of ALS reside mostly in one
bundle, namely the CST.

@@

To test the performance of SGL \deleted{with tractometry data} in a continuous
regression task, we focus here on the prediction of biological age in three
datasets named WH, HBN, and Cam-CAN (see the methods section for a description of each dataset). Prediction of ``brain
age'' is a commonly undertaken task in neuroimaging machine learning, in part
because these predictions, and deviations therefrom, may be diagnostic of
overall brain health (for a review, see Cole et al.~\cite{Cole2019-rz}).
However, as Nelson et al.~\cite{nelson2019biomarkers} have observed, aging
biomarkers are subject to unique challenges and tend to be noncausative. Our
interest in aging here relies on its utility as a methodological benchmark.
Biological age operates on a natural scale, with meaningful and easily
understood units, and it's popularity as a machine learning target makes it
valuable for comparisons with other studies.

@@

To evaluate the fit of the model, we used a nested cross-validation
procedure. In this procedure, batches of subjects are held out. For each
batch (or fold), the model is fully fit without this data. Then, once the
parameters are fixed, the model is applied to predict the ages of held out
subjects based on the linear coeffiecients. This scheme automatically finds
the right level of sparsity and fits the coefficients to the ill-posed linear
model, while guarding against overfitting. SGL accurately predicts the age of
the subjects in this procedure, with a median absolute error of {\whMae},
{\hbnMae}, and {\camcanMae} years for the WH, HBN, and Cam-CAN datasets,
respectively and coefficients of determination $R^2 = $ {\whRsq} , {\hbnRsq},
and {\camcanRsq}, respectively (see \cref{fig:age-results}, top panels). The
predictions for Cam-CAN are competitive with a recent state of the art
prediction \cite{mcpherson2020single}, which used streamline density to
estimate the brain's structural connectivity and achieved $R^2 = 0.63$. The
median absolute errors are also lower than the results of a recent study that
predicted age in a large sample that included the Cam-CAN data, and was based
on diffusion MRI features \cite{Richard2018-ux}.

@@

Specifically, we demonstrated that the algorithm correctly detects the fact
that ALS, which is a disease of lower motor neurons, is localized to the
cortico-spinal tract. This recapitulates the results of previous analysis of
these same data, using a bundle-of-interest approach
\cite{sarica2017corticospinal}.
\added{As with the original study, it is unclear whether our strong predictive performance based on image properties that would be visible in patient scans or whether it is identifying a new subclinical disease biomarker.}

In contrast, for the analysis of biological
age, the coefficients identified by the algorithm are very widely distributed
across many parts of the white matter, mirroring previous results that show a
large and continuous distribution of life-span changes in white matter
properties \cite{yeatman2014lifespan}.
\added{We present age quintile bundle profiles and SGL \protect$\beta$ coefficients for all four datasets in the supplementary material. For age regression, substantial differences between the datasets preclude generalization across all three:}
\begin{enumerate*}[%
    label=(\roman*),%
    before=\unskip{: },%
    itemjoin={{, }},%
    itemjoin*={{, and }}]
    \item \added{different acquisition parameters, which can be challenging to harmonize \mbox{\cite{pinto2020harmonization}}}
    \item \added{different diffusion models, with DTI for the WH dataset and DKI for the HBN and Cam-CAN datasets}
    \item \added{different age ranges and distributions (which is evident in the figure legends for \cref{fig:age-results}), with HBN being a developmental dataset, while WH and Cam-CAN are lifespan maturation datasets}
    \item \added{different anatomical extents, with the WH streamlines truncated to remain with the bundle's bounding regions of interest (the default behavior in the legacy \texttt{mAFQ}) and the HBN and Cam-CAN streamlines allowed to retain their full extent from whole-brain tractography (the default behavior in \texttt{pyAFQ}).}
\end{enumerate*}
\added{In a biological characteristic that has widespread effects in the brain, validity is difficult to assess \mbox{\cite{lermausabiaga2019replication}}, so it is both unsurprising and vexing that we can do so well at predicting within each dataset, and yet do poorly in interpreting informative features across instruments.}

@@

\added{Another limitation of our approach is in its computational execution time,}
\added{which, in our experience, is roughly five times slower than simpler models}
\added{like the lasso. This slow-down is likely to be insignificant for ``production''}
\added{cycles in which researchers report their findings, but may be prohibitive}
\added{for ``development'' cycles in which researchers repeatedly train models while}
\added{adjusting their analysis pipeline. In practice, we circumvent this limitation}
\added{by developing our analysis infrastructure with fast, simple models, and substituting}
\added{slower, more performant models afterward.}

@@

The method is packaged as open-source software called AFQ-Insight that is
openly available at \url{https://github.com/richford/AFQ-Insight}, and
provides a clear API to allow for extensions of the method. The sofware
integrates within a broader automated fiber quantification software
ecosystem: AFQ \cite{yeatman2012tract} and pyAFQ \cite{pyAFQ, kruper2021evaluating}, which extract
\replaced{tract profile}{tractometry} data from raw and processed dMRI datasets, as well as
AFQ-Browser, which visualizes \replaced{tract profiles}{tractometry} data and facilitates sharing of the
results of dMRI studies \cite{yeatman2018browser}. To facilitate
reproducibility and ease use of the software, the results presented in this
paper are also provided in
\url{https://github.com/richford/afq-insight-paper} as a series of Jupyter
notebooks \cite{kluyver2016jupyter}.

@@

\item \added{{\it MRtrix3 Reconstruction}}

\added{Reconstruction was performed using \emph{QSIprep} 0.12.1. Multi-tissue fiber response functions were estimated using the dhollander algorithm. FODs were estimated via constrained spherical deconvolution (CSD, \mbox{\cite{originalcsd, tournier2008csd}}) using an unsupervised multi-tissue method \mbox{\cite{dhollander2019response, dhollander2016unsupervised}}. Reconstruction was done using MRtrix3 \mbox{\cite{mrtrix3}}. FODs were intensity-normalized using mtnormalize \mbox{\cite{mtnormalize}}.}
\end{itemize}

@@

\item Diffusion MRI data from the Cambridge Centre for Ageing and
Neuroscience (Cam-CAN) ``CC700'' dataset
\cite{shafto2014cambridge,taylor2017cambridge}, containing data from 640
subjects with ages 18-88. These data were measured on a 3T Siemens TIM Trio
system and written informed consent was obtained from each participant. Voxel
resolution was \num{2x2x2}$\text{mm}^3$ with 30 non-colinear directions
measured for each of $b=1000$ $\text{s} / \text{mm}^2$ and $b=2000$
$\text{s} / \text{mm}^2$. The diffusion weighted images were acquired
with a twice refocused spin-echo sequence and the same preprocessing \added{and reconstruction} pipeline
used for the HBN dataset was applied to this data. We will refer to this
dataset as Cam-CAN.

@@

Data from the ALS and WH studies was processed in a similar manner,
using the Matlab Automated Fiber Quantification toolbox
(\texttt{mAFQ}\added{, version 1.1 for WH and version 1.2 for ALS})
\cite{yeatman2012tract}: streamlines representing fascicles of white
matter tracts were generated using a determinstic tractography algorithm
that follows the prinicpal diffusion direction of the diffusion tensor
in each voxel (STT) \cite{basser2000vivo}. \replaced{Eighteen major}{Major}
tracts\added{, which are enumerated in the supplemental material,} were identified
using multiple criteria: streamlines are selected as candidates for
inclusion in a bundle of streamlines that represents a tract if they
pass through known inclusion ROIs and do not pass through exclusion
ROIs \cite{wakana2007reproducibility}. In addition, a probabilistic
atlas is used to exclude streamlines which are unlikely to be part of
a tract \cite{Hua2008-sh}. Each streamline is resampled to 100 nodes
and the robust mean at each location is calculated by estimating the 3D
covariance of the location of each node and excluding streamlines that
are more than 5 standard deviations from the mean location in any node.
Finally, a bundle profile of tissue properties in each bundle was created
by interpolating the value of MRI maps of these tissue properties to the
location of the nodes of the resampled streamlines designated to each
bundle. In each of 100 nodes, the values are summed across streamlines,
weighting the contribution of each streamline by the inverse of the
\replaced{mahalanobis}{mahalnobis} distance of the node from the average of
that node across streamlines. This means that streamlines that are more
representative of the tract contribute more to the bundle profile, relative
to streamlines that are on the edge of the tract.

Data from the HBN and Cam-CAN studies were processed using the updated Python
Automated Fiber Quantification toolbox (\texttt{pyAFQ}; \cite{pyAFQ}). In
addition to demonstrating the our analysis pipeline is robust to changes in
tractometry software, the use of the updated \texttt{pyAFQ} capitalized upon
the following improvements over the legacy Matlab version
\begin{enumerate*}[%
    label=(\roman*),%
    before=\unskip{: },%
    itemjoin={{, }},%
    itemjoin*={{, and }}]
    \item the ability to ingest data provided in the BIDS format
    \cite{gorgolewski2016brain}
    \item the calculation of diffusion kurtosis imaging (DKI
    \cite{jensen2005diffusion}) metrics
\end{enumerate*}
We will refer to the \texttt{mAFQ} and \texttt{pyAFQ} pipeline collectively
as \texttt{AFQ}.
